Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2022-03-07 paper/paper.tex
WIP

We asked i) whether this mechanism could \emph{replace} the non-biological
competition, C, (making necessary its extension to the dorsal-ventral retinal
axis) and ii) if cast in terms of receptor-ligand interactions, could the
strong ligand expression assumption be relaxed?

@@

Taken alone, the EphA forward interaction between one RGC branch, $b$, and
another, $k$, will cause a repulsion if $k$ originated more nasally than
$b$. A branch originating on the \emph{temporal} side of $b$ will cause no
repulsion of $b$. However, although $k$ does not induce repulsion in $b$, $b$
should induce a repulsion in $k$ and thus, it could be argued that only a
single ephrin-A ligand gradient should be required in the retina to ensure
that repulsive signals are transmitted between axons whose cell bodies
originte far apart in the retina. However, there is an asymmetry in the
interaction. Consider a cluster of growth cones, $k$, originating from the
nasal retina near to a single cone, $b$, from the temporal retina and assume
that $r_b/r_k>1.1$. The single cone is affected by the sum of 3 interactions
and has a movement of 3 units added to its movement vector (in the temporal
direction). Now reverse the originating locations so that the cluster, $k$,
originates temporally and the individual cone, $b$, comes from the nasal
retina. Now $r_k/r_b>1.1$ and each member of the cluster experiences an
addition of just one unit to its movement vector and the cones move apart at a
speed that is only $1/3$ of that for the original origins.

@@

where $r_{\!_J}$ is the receptor-ligand interaction radius, $l_{i,k}$ (the
expression of ligand $i$ on branch $k$) is the $i^{\mathrm{th}}$ element of
$\mathbf{l}_k$ and $r_{i,b}$ is the $i^{\mathrm{th}}$ element of
$\mathbf{r}_b$. $F_i$, defined in Eq.\,\ref{e:G}, is -1 if the $r_{i}\,l_{i}$
interaction signals repulsion and +1 if it signals attraction. $s$ is a
positive signal threshold whose value should be around 4.7727; if higher, then
branches whose retinal origin is close will not experience a competitive
interaction. If $s\leq 4.7727$, then \emph{any} pair of branches will compete as they
approch each other. With the constraint $s>0$ comes the assumption that
attractive interactions will have no effect on the competitive movement
vector---a receptor-ligand pair that signals attraction cannot cause
$Q_{\!_J}$ to take the value 1. $Q_{\!_J}$ is set to 1 only if at least one
repulsive signal exceeds the threshold, $s$ and the distance from branch
$b$ to branch $k$ is smaller than $2 r_{\!_J}$.

@@

% Now the simulation.
Adopting retinal expressions given in Eqs.\,\ref{e:retrcpt}
and \ref{e:retligexp}, we performed a simulated annealing optimization to find
the best parameters for the $\mathbf{GJ}$ model.  With $m_{\!_G}$ fixed at the
value 0.0012, we allowed the threshold parameter $s$ to vary along with
$m_{\!_J}$ and $r_{\!_J}$ . The parameters best able to recover the wildtype
pattern were $s=4.732$, $m_{\!_J} = 0.247$ and $r_{\!_J} = 0.079$. Note that
the discovered value for $s$ is less than 4.7727 meaning that for \emph{any
pair of branches}, either $r_0\times l_0$ or $r_2\times l_2$ will exceed $s$.

The
simulation of the system with these parameters is shown in Fig.\,\ref{f:GJ}
and Fig.\,\ref{f:GJsurg}.

@@

@@

% ./scripts/fig_GC_surgical.sh
\begin{figure}
\includegraphics[width=\linewidth]{./images/fig_GC_surgical.png}
\caption{$\mathbf{GC}$ model, surgical manipulations. Demonstrating that this
model reproduces these manipulations.}
\label{f:GCsurg}
\end{figure}

\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_GC_best_1_reber_exit_true_steps_1500.png}
\caption{$\mathbf{GC}$ model, Reber experiment.}
\label{f:GCreber}
\end{figure}

>>>>

We asked i) whether this mechanism could \emph{replace} the non-biological
competition, $\mathbf{C}_b$, and ii) if cast in terms of receptor-ligand
interactions, could the strong ligand expression assumption be relaxed?

@@

Taken alone, the EphA forward interaction between one RGC branch, $b$, and
another, $k$, will cause a repulsion if $k$ originated more nasally than
$b$. A branch originating on the \emph{temporal} side of $b$ will cause no
repulsion of $b$. However, although $k$ does not induce repulsion in $b$, $b$
should induce a repulsion in $k$ and thus, it could be argued that only a
single ephrin-A ligand gradient should be required in the retina to ensure
that repulsive signals are transmitted between axons whose cell bodies
originate far apart in the retina. However, there is an asymmetry in the
interaction. Consider a cluster of growth cones, $k$, originating from the
nasal retina near to a single cone, $b$, from the temporal retina and assume
that $r_b/r_k>1.1$. The single cone is affected by the sum of 3 interactions
and has a movement of 3 units added to its movement vector (in the temporal
direction). Now reverse the originating locations so that the cluster, $k$,
originates temporally and the individual cone, $b$, comes from the nasal
retina. Now $r_k/r_b>1.1$ and each member of the cluster experiences an
addition of just one unit to its movement vector and the cones move apart at a
speed that is only $1/3$ of that for the original origins.

@@

where $r_{\!_J}$ is the receptor-ligand interaction radius, $l_{i,k}$ (the
expression of ligand $i$ on branch $k$) is the $i^{\mathrm{th}}$ element of
$\mathbf{l}_k$ and $r_{i,b}$ is the $i^{\mathrm{th}}$ element of
$\mathbf{r}_b$. $F_i$, defined in Eq.\,\ref{e:G}, is -1 if the $r_{i}\,l_{i}$
interaction signals repulsion and +1 if it signals attraction. $s$ is a
positive signal threshold whose value should be around 4.7727; if higher, then
branches whose retinal origin is close will not experience a competitive
interaction. If $s\leq 4.7727$, then \emph{any} pair of branches will compete as they
approach each other. With the constraint $s>0$ comes the assumption that
attractive interactions will have no effect on the competitive movement
vector---a receptor-ligand pair that signals attraction cannot cause
$Q_{\!_J}$ to take the value 1. $Q_{\!_J}$ is set to 1 only if at least one
repulsive signal exceeds the threshold, $s$ and the distance from branch
$b$ to branch $k$ is smaller than $2 r_{\!_J}$.

@@

% Now the simulation.
Adopting retinal expressions given in Eqs.\,\ref{e:retrcpt}
and \ref{e:retligexp}, we performed a simulated annealing optimization to find
the best parameters for the $\mathbf{GJ}$ model.  With $m_{\!_G}$ fixed at the
value 0.0012, we allowed the threshold parameter $s$ to vary along with
$m_{\!_J}$ and $r_{\!_J}$. The parameters best able to recover the wildtype
pattern were $s=4.732$, $m_{\!_J} = 0.247$ and $r_{\!_J} = 0.079$, with the
wildtype result shown in Fig.\,\ref{f:GJ}. Note that the discovered value for
$s$ is \emph{less} than 4.7727 meaning that for any pair of branches, either
$r_0\times l_0$ or $r_2\times l_2$ will exceed $s$ and branches all
compete. With $s < 4.7727$ the $\mathbf{GJ}$ model is equivalent to the
$\mathbf{GC}$ model. However, the parameters found for the $\mathbf{GJ}$ model
did lead to better results for surgical manipulations, as shown in
Fig.\,\ref{f:GJsurg}. Values for the pattern metric, $\epsilon$ are
consistently lower than those for the $\mathbf{GC}$ model shown in
Fig.\,\ref{f:GCsurg}. Given that the \emph{systems} are equivalent, the
differences must have arisen due to the optimization process, which must have
found alternative `good' parameters. We verified this by copying the
parameters found for the $\mathbf{GJ}$ model into the $\mathbf{GC}$ model,
setting $m_G = 0.0012$, $r_c = 0.079$ and $m_c =
0.247$. Fig.\,\ref{f:GClikeGJsurg} shows the result for surgical
manipulations. Their equivalence to Figs.~\ref{f:GCsurg} is evident by
inspection. Nevertheless, we retain the receptor-ligand signalling of the
$\mathbf{GJ}$ model as this maintains the ability of the model to respond to
the genetic manipulations described in the next section.

@@

In the early 2000s, a series of genetic experiments added to the set of
manipulations that a model of the retinotectal system would need to
explain. First, \citet{brown_topographic_2000} showed that a selective RGC
knock-in of EphA3 receptors led to an interesting modulation of the maps found
in the tectum. For the strongest knock-in (a homozygote manipulation) two maps
were formed, with knock-in cells finding termination zones more rostral than
their wildtype retinal neighbours. By enhancing the quantity of EphA receptors
expressed by half of all retino-tectal axons, they demonstrated the importance
of retinal receptor expression for the arrangement of the tectal map. In the
weaker, heterozygote manipulation, the maps were still separated, but the
separation `collapsed' towards the rostral end of the tectum
(see \citet{brown_topographic_2000}, Fig\,5). This led to a second study
showing that with an additional knock-down of EphA4 receptors in \emph{all}
RGCs, the retinal origin of the collapse point was shifted temporally and all
heterozygote EphA3 knock-in cells were pushed even further rostrally
(\citet{reber_relative_2004}, their Fig.\,3e). Does the $\mathbf{GJ}$ model
reproduce the results of Brown et al and Reber et al?

An assumption made in these papers is that all EphA receptor subtypes
contribute to a summed EphA expression level, $\sum\mathrm{EphA}$, which
increases from a low level at the nasal side of the retina, to a high level at
the temporal side. The precise contribution of each subtype to the overall
level is assumed not to be important. In our model, receptor $r_0$ increases
in the retina in a nasal to temporal direction (Fig\,\ref{f:expressions}B) and
so we will manipulate $r_0$ to recreate the Brown et al.~and Reber et
al.~experiments. In half of the RGCs, we knock-in EphA3 by adding to the value
of $r_0$ in these cells either 0.6 (heterzygote EphA3 ki/+) or 3.0 (homozygote
EphA3 ki/ki). To add the additional knock-down of EphA4
in \citet{reber_relative_2004}, we subtract 0.35 from $r_0$ in \emph{all}
RGCs.

@@

% ./scripts/fig_GC_surgical.sh
\begin{figure}
\includegraphics[width=\linewidth]{./images/fig_GC_surgical.png}
\caption{$\mathbf{GC}$ model, surgical manipulations. Demonstrating that this
model performs better than the chemoaffinity-only model.}
\label{f:GCsurg}
\end{figure}

% ./scripts/fig_GClikeGJ_surgical.sh
\begin{figure}
\includegraphics[width=\linewidth]{./images/fig_GClikeGJ_surgical.png}
\caption{$\mathbf{GC}$ model with parameters found from \emph{GJ}
optimization ( $m_G = 0.0012$, $r_c = 0.079$ and $m_c = 0.247$). Results are
equivalent to those for the $\mathbf{GJ}$ model in Fig.\,\ref{f:GJsurg}.}
\label{f:GClikeGJsurg}
\end{figure}

\begin{figure}
\includegraphics[width=\linewidth]{./images/EphA_manipulations_GClikeGJ.png}
\caption{$\mathbf{GC}$ model with parameters found from \emph{GJ}
optimization, genetic manipulations. As for the surgical manipulations, the
results are similar to those of the $\mathbf{GJ}$ model with the equivalent
parameters. However, there are subtle differences such as XXX.}
\label{f:GClikeGJreber}
\end{figure}
